Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/appendix.tex
laborious account of functoriality and sq preservation for Kleisli operations

In CBPV models, all the type constructors are interpreted as functors:
\begin{enumerate}
\item $\to : \op\calV \times \calE \to \calE$
\item $\times : \calV \times \calV \to \calV$
\item $F : \calV \to \calE$
\item $U : \calE \to \calV$
\end{enumerate}
That is, they all have functorial actions on \emph{pure} morphisms of
value types and \emph{linear} morphisms of computation types.
%
We use these functorial actions extensively in the construction of
casts and their corresponding perturbations. But when defining
downcasts of value types and upcasts of computation types, we
additionally need a second functorial action of these categories:
functoriality in \emph{impure} morphisms of value types and
\emph{non-linear} morphisms of computation types. These notions of
morphism are given by the \emph{Kleisli} categories $\calVk$ and
$\calEk$ which have value types and computation types as objects but
morphisms are defined as
\[ \calVk(A,A') = \calE(F A, FA')\]
\[ \calEk(B,B') = \calV(U B, U B')\]
with composition given by composition in $\calE/\calV$.  That is we
need to define a second functorial action, that agrees with the above
on objects for these Kleisli categories:
\begin{enumerate}
\item $\tok : \op\calVk \square \calEk \to \calEk$
\item $\timesk : \calVk \square \calVk \to \calVk$
\item $\Fk : \calVk \to \calEk$
\item $\Uk : \calEk \to \calVk$
\end{enumerate}
Note that rather than the product of categories we use the ``funny
tensor product'' $\square$. This is because the action on
impure/non-linear morphisms for $\tok/\timesk$ do not satisfy ``joint
functoriality'' but instead only ``separate functoriality'', meaning
we give rather than an action on morphisms in both categories
simultaneously instead an action on each argument categories morphisms
with the object in the other category fixed. The existence of these
functorial actions for $\tok$ and $\timesk$ is reliant on the
\emph{strength} of the adjunction. We describe them using the internal
language of CBPV in order to more easily verify their
existence/functoriality:
\begin{enumerate}
\item For $\tok$ we define for $\phi : \calE(F A,F A')$ and $B \in \calE$ the morphism $\phi \tok B : \calV(U(A' \to B),U(A\to B))$ as
  \[ t:U(A'\to B) \vdash \phi \tok B = \{ \lambda x. x' \leftarrow \phi\,[\ret x]; ! t x'\} : U(A \to B) \]
  and for $A \in \calV$ and $f : \calV(UB,UB')$ we define $A \tok f : \calV(U(A \to B),U(A\to B'))$ as
  \[ t : U(A \to B) \vdash A \tok f = \{ \lambda x. !f[\{ ! t x \}]\} \]
\item For $\timesk$ we define for $\phi : \calE(F A_1,FA_2)$ and $A' \in \calV$ the morphism $\phi \timesk A_2$ as
  \[ \bullet : F(A_1\times A_2) \vdash \phi \timesk A_2 = (x_1,x_2) \leftarrow \bullet; x_1' \leftarrow \phi[\ret x_1]; \ret (x_1',x_2) : F(A_1'\times A_2)\]
  and $A_1 \timesk \phi$ is defined symmetrically.
\item For $\Uk$ we need to define for $f : \calV(UB,UB')$ a morphism $\Uk f : \calE(FUB,FUB')$. This is simply given by the functorial action of $F$: $\Uk f = F(f)$
\item Similarly $\Fk \phi = U\phi$
\end{enumerate}

Functoriality in each argument is easily established, meaning for
example for the function type is functorial in each argument:
\begin{enumerate}
\item $(\phi \circ \phi') \tok B = (\phi' \tok B) \circ (\phi \tok B)$
\item $\id \tok B = \id$
\item $A \tok (f \circ f') = (A \tok f) \circ (A \tok f)$
\item $A \tok \id = \id$
\end{enumerate}

Finally, note that all of these constructions lift to squares in a
double CBPV model since the squares themselves form a CBPV model and
the projection functions preserve CBPV structure. For instance, given a square
$\alpha : \phi \ltdyn_{F c_o}^{F c_i} \phi'$ and a horizontal morphism $d : B \rel B'$ of appropriate type, we get a square
\[ \alpha \tok d : \phi \tok B \ltdyn_{U(c_o \to d)}^{U(c_i \to d)} \phi' \tok B' \]

>>>>

We now define the Kleisli actions of the type constructors. These are the following
\begin{enumerate}
\item $- \tok B : \op\PreDom_k \to \ErrDom_k$
\item $A \tok - : \ErrDom_k \to \ErrDom_k$
\item $- \timesk A_2 : \PreDom_k \to \PreDom_k$
\item $A_1 \timesk - : \PreDom_k \to \PreDom_k$
\end{enumerate}
While these can be defined in any CBPV model, for simplicity we just
give their definition explicitly for predomains and error domains:
\begin{definition}{~}
  Given $\phi : \li A_i \multimap \li A_o$, we define $\phi \tok B : U(A_o \to B) \to U(A_i \to B)$ as
  \[ (\phi \tok B)(f)(x) = f^\dagger(\phi(\eta(x))) \]

  This is functorial in that $\id \tok B = \id$ and $(\phi \circ
  \phi') \tok B = (\phi' \tok B) \circ (\phi \tok B)$.

  Further, this preserves squares in that if $\phi \ltsq{\li c_i}{\li c_o} \phi'$ then $\phi \tok B \ltsq{U(c_o \to d)}{U(c_i \to d)} \phi' \tok B'$.
\end{definition}
\begin{proof}
  {~}
  \begin{itemize}
  \item For identity
    \[ (\id \tok B)(f)(x) = f^\dagger(\eta(x)) = f(x) \]

  \item 
    For composition, expanding definitions it suffices to show
    \[ f^\dagger(\phi(\phi'(\eta(x)))) = ((\phi \tok B)(f))^\dagger(\phi'(\eta(x))) \]
    Which follows if we can show
    \[ ((\phi \tok B)(f))^\dagger = f^\dagger \circ \phi \]
    By the freeness of $\li -$ it suffices to show it for inputs of the form $\eta y$:
    \[ ((\phi \tok B)(f))^\dagger(\eta y) = ((\phi \tok B)(f))(y) = f^\dagger(\phi(\eta y))\]
  \item For squares, we assume $f \binrel{U(c_o\to d)} f'$ and $x
    \binrel {c_i} x'$ and we need to show $f^\dagger(\phi(x)) \binrel
    d (f')^\dagger(\phi'(x))$. This follows easily since $f^\dagger
    \ltsq{\li c_o}{d} f^{\prime\dagger}$.
  \end{itemize}
\end{proof}

\begin{definition}
  Given $f : UB \to UB'$, define $A \tok f : U(A \to B) \to U(A \to B')$ as
    \[ (A \tok f)(g)(x) = f(g(x)) \]

  This is functorial in that $A \tok \id = \id$ and $A \tok (f \circ
  f') = (A \tok f) \circ (A \tok f')$.

  Further this preserves squares in that if $f \ltsq{Ud_i}{Ud_o} f'$
  then $A \tok f \ltsq{U(c \to d_i)}{U(c \to d_o)} A' \tok f'$
\end{definition}
\begin{proof}
  \begin{itemize}
  \item For identity
    \[ (A \tok \id)(g)(x) = g(x) \]
  \item 
    for composition
    \[  (A \tok (f \circ f'))(g)(x) = f(f'(g(x))) = (A\tok f)((A \tok f')(g))(x) \]
  \item 
   for squares, given $g \binrel{U(c \to d_i)} g'$ and $x \binrel{c}
   x'$ we have $f(g(x)) \binrel{d_o} f'(g'(x'))$.
  \end{itemize}
\end{proof}

\begin{definition}[Kleisli Actions of $\times$]
  Given $\phi_1 : \li A_1 \multimap \li A_1'$ we define $\phi_1 \timesk A_2 : \li (A_1 \times A_2) \multimap (A_1' \times A_2)$ as the unique extension of the following to a homomorphism:
  \[ (\phi_1 \timesk A_2)(\eta (x_1,x_2)) = (\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1(\eta(x_1))) \]
  and similarly define
  \[ (A_1 \timesk \phi_2)(\eta (x_1,x_2)) = (\lambda x_2'. \eta(x_1,x_2'))^\dagger(\phi_2(\eta(x_2)))\]
\end{definition}
\begin{proof}
  We show cases only for $-\timesk A_2$ as the other is symmetric.
  \begin{itemize}
  \item For identity it is sufficient to consider inputs of the form $\eta(x_1,x_2)$, then indeed
    \[ (\id \timesk A_2)(\eta(x_1,x_2)) = (\lambda x_1'. \eta(x_1',x_2))^\dagger(\eta(x_1)) = \eta(x_1,x_2) \]
  \item For composition, again for pure inputs this reduces to
    \[ (\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1'(\phi_1(\eta(x_1))))
    = (\phi_1'\timesk A_2)((\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1(\eta(x_1)))) \]
    So it suffices to show
    \[ (\lambda x_1'. \eta(x_1',x_2))^\dagger \circ \phi_1' = (\phi_1'\timesk A_2) \circ (\lambda x_1'. \eta(x_1',x_2))^\dagger \]
    Since both sides are homomorphisms out of the free error domain by freeness it suffices to show they are equal for inputs of the form $\eta(x_1')$:
    \[  (\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1'(\eta(x_1')))
    = (\phi_1'\timesk A_2)(\eta(x_1',x_2))
    = (\phi_1'\timesk A_2)((\lambda x_1'. \eta(x_1',x_2))^\dagger(\eta(x_1')))
    \]

  \item Finally, for squares, we need to show if $\phi_l \ltsq{\li
    c_1}{\li c_1'} \phi_r$ then $\phi_l \timesk A_{2l} \ltsq{\li (c_1
    \times c_2)}{\li (c_1' \times c_2)} \phi_r \timesk A_{2r}$.

    By the universal property of $\li-$ on relations, it suffices to
    assume $x_{1l} \binrel{c_1} x_{1r}$ and $x_{2l} \binrel{c_2} x_{2r}$ and prove
    \[ (\phi_l \timesk A_{2l})(\eta(x_{1l},x_{2l})) \binrel{(\li (c_1' \times c_2))} (\phi_r \timesk A_{2r})(\eta(x_{1r},x_{2r})) \]
    expanding definitions this is
    \[ (\lambda x_{1l}'. \eta(x_{1l}',x_{2l}))^\dagger\phi_l(\eta(x_{1l}))
    \binrel{(\li (c_1' \times c_2))}
    (\lambda x_{1r}'. \eta(x_{1r}',x_{2r}))^\dagger\phi_r(\eta(x_{1r}))
    \]
    By our assumptions we have $\phi_l(\eta(x_{1l}) \binrel {\li c_1'} \phi_r(\eta(x_{1r})$ so since $-^\dagger$ preserves squares it suffices to show
    \[ (\lambda x_{1l}'. \eta(x_{1l}',x_{2l})) \ltsq{\li c_1}{\li (c_1' \times c_2)} (\lambda x_{1r}'. \eta(x_{1r}',x_{2r})) \]
    To show this assume $x_{1l}' \binrel{c_1'} x_{1r}'$. Then we have
    \[ \eta(x_{1l}',x_{2l}) \binrel{\li (c_1' \times c_2)} \eta(x_{1r'},x_{2l})  \]
    holds.
  \end{itemize}
\end{proof}
